### 2020-06-18 report/sections/abstract.tex
Report: final proofread

\noindent
Multiparty Session Types (MPST)
is a typing discipline for concurrent
processes that communicate via
message-passing. MPST theory can ensure
communication safety properties and
protocol conformance.
Existing work in session-typed
web development over WebSocket transport
is incompatible with modern web
programming practices and are limited
to supporting communication protocols
that implement the server-centric network
topology.

We address limitations in the current 
state-of-the-art by:
\textbf{(1)} implementing \codegen, a code generation
toolchain for session-typed web development
over WebSocket transport using TypeScript;
and \textbf{(2)} presenting \newtheory, a new
multiparty session type theory that supports
routed communications.

\codegen provides
developers with TypeScript APIs generated
from a communication protocol specification 
based on multiparty session type theory.
Our work is compatible with modern web programming
industrial practices. The generated APIs
build upon TypeScript concurrency primitives,
complement the event-driven style of programming
in full-stack web development,
and are compatible
with the Node.js runtime and the React.js framework
for server-side and browser-side endpoints respectively.
We evaluate the expressiveness of \codegen for modern
web programming
using case studies of protocols found in web-services,
and analyse the overhead of our implementation through 
running micro-benchmarks against a baseline implementation.

>>>>

\noindent
Modern web programming
involves coordinating interactions
between web browser clients and a web server.
Typically, the interactions are
informally described, making
it difficult to verify communication
correctness in web-based distributed systems.

Multiparty Session Types (MPST)
are a typing discipline for concurrent
processes that communicate via
message passing. MPST theory can ensure
communication safety properties and
protocol conformance.
Existing work in session-typed
web development over WebSocket transport
is incompatible with modern web
programming practices and is limited
to supporting communication protocols
that implement the server-centric network
topology.

We address limitations in the current 
state of the art by:
\textbf{(1)} implementing \codegen, a code generation
toolchain for session-typed web development
over WebSocket transport using TypeScript;
and \textbf{(2)} presenting \newtheory, a new
multiparty session type theory that supports
routed communications.

\codegen provides
developers with TypeScript APIs generated
from a communication protocol specification 
based on multiparty session type theory.
Our work is compatible with modern web programming
industrial practices. The generated APIs
build upon TypeScript concurrency practices,
complement the event-driven style of programming
in full-stack web development,
and are compatible
with the Node.js runtime for server-side endpoints
and the React.js framework
for browser-side endpoints.
We evaluate the expressiveness of \codegen for modern
web programming
using case studies of protocols found in web services,
and analyse the performance overhead
through 
running benchmarks against a baseline implementation.

### 2020-06-18 report/sections/background/related.tex
Report: final proofread

The protocol specification language is 
a component of the broader 
\textit{Scribble toolchain} initiated by 
Honda et al. in \cite{Scribble}, 
through which the toolchain also facilitates 
the development of 
\textit{endpoint applications} that 
conform to user-specified protocols.

A Scribble global protocol can be projected
to a role, or \textit{endpoint},
to obtain a local protocol which represents
the global protocol viewed from the perspective
of the endpoint.
This allows the endpoint to \textit{verify} their
implementation against their local protocol
for conformance, independent of other endpoints.
The communication safety guarantees from MPST
theory also apply here: if the implementation
for each endpoint is verified against their
local protocol, the distributed system as
a whole will conform to the global protocol.

@@

Developers can use the EFSM to verify their implementation
for protocol conformance with respect to an endpoint
in the global protocol.
Endpoint API generation is a common approach
\cite{HuJava,Hybrid2016,Python2017,PureScript2019,LinearDecomp}:
these proposals generate APIs in the target language
which the developer
can use to implement their endpoint application that guarantees
the following properties:

@@

\subsubsection{API Generation in PureScript \cite{PureScript2019}}
King et al. presented an approach
for integrating \textit{multiparty} session types 
into web development using 
the PureScript language \cite{PureScript},
PureScript is a strongly typed functional language
that compiles to JavaScript.
This proposal takes advantage of its expressive type system to 
provide \textit{static} guarantees for protocol conformance.
We outline the main components of their EFSM encoding:

\subparagraph{Actions as type classes} 
The semantics of the state transition function 
in the EFSM express that a (state, action) tuple
\textit{uniquely defines} a successor state. 
These semantics can be expressed by 
\textit{multi-parameter type classes (MPTC) 
with functional dependencies}: 
\lstonelineps{class Send r s t a | s -> t r a} 
defines \texttt{Send} as a MPTC 
parameterised by recipient \texttt{r}, 
current state \texttt{s}, successor state \texttt{t} 
and payload type \texttt{a}; 
\texttt{s -> t r a} expresses the functional dependency that, 
for an instance of this type class, 
the current state uniquely determines the
successor state, the recipient and payload type. 
These type classes are independent of the EFSM.

@@

\subsubsection{Session-Typed GUI Programming 
using Links \cite{MVU2020}}
Fowler presented the first formal integration
of session typing and GUI programming .
His work formalised the \textit{Model-View-Update} (MVU)
architectural pattern for GUI development.
An MVU application features a \textit{model} encapsulating application
state, a \textit{view function} rendering the state on the 
user interface, and an
\textit{update function} handling
\textit{messages} produced by the
rendered model to produce a new model.
Fowler introduces the concept of model types
to express type dependencies between these components:
a \emph{model type} uniquely defines a \textit{view} function,
set of \textit{messages} and \textit{update function} --
rather than producing a new model, the update function 
defines valid transitions to other model types.
This approach guarantees channel linearity for GUI programming:
the update function transitions to a successor model, which only
renders an interface with the set of messages unique to that model,
so channel actions from the predecessor model are not rendered,
which makes reuse impossible.
The core MVU calculus is also extended with \textit{commands}
and \textit{linearity} required to express communication
actions in session type theory.

Fowler's proposal \cite{MVU2020} is implemented using
the \textit{Links} web programming language
\cite{LINKS}. Links is a strict, typed, functional language
that can be used to write full-stack web applications;
and similar to PureScript, it is compiled to JavaScript.
As it has a linear type system, session types can be
implemented using native language features, since the
language can monitor linear channel usage.
Links also supports error handling for session types,
as introduced in \cite{Exceptional}; 
given side-effects must be made explicit
in functional programming,
Fowler et al. \cite{Exceptional} introduces
a \textit{session cancellation} operator to ensure
that channel resources are closed after handling
side-effects, such as errors.

>>>>

The protocol specification language is 
a component of the broader 
\textit{Scribble toolchain} initiated by 
Honda et al. \cite{Scribble}, 
through which the toolchain also facilitates 
the development of 
\textit{endpoint applications} that 
conform to user-specified protocols.

A Scribble global protocol can be projected
to a role, or \textit{endpoint},
to obtain a \textit{local protocol} 
which represents
the global protocol viewed from the perspective
of the endpoint.
This allows the endpoint to \textit{verify} their
implementation against their local protocol
for conformance, independent of other endpoints.
The communication safety guarantees from MPST
theory also apply here: if the implementation
for each endpoint is verified against their
local protocol, the distributed system as
a whole will conform to the global protocol.

@@

We also inherit the characteristics of EFSMs derived
from well-formed protocols, as presented in \cite{Hybrid2016}:

\begin{itemize}
\item 
There is \textit{exactly one} initial state 
(\texttt{S1} in \cref{fig:bgadderfsm}).

An initial state has no incoming transitions.
Formally, the state identifier of the initial state
is not in the range of $\delta$.

\item 
There is \textit{at most one} terminal state 
(\texttt{S2} in \cref{fig:bgadderfsm}).

A terminal state has no outgoing transitions.
Formally, the state identifier of the terminal state
is not in the domain of $\delta$.

\item
Every non-terminal state is either a \textit{send}
state or a \textit{receive} state.

A send state only has send actions in its
outgoing transitions. \texttt{S3} and \texttt{S4}
are the send states in \cref{fig:bgadderfsm}.

A receive state only has receive actions in its
outgoing transitions. \texttt{S1} is the only
receive state in \cref{fig:bgadderfsm}.
\end{itemize}

Developers can use the EFSM to verify their implementation
for protocol conformance with respect to an endpoint
in the global protocol.
Endpoint API generation is a common approach
\cite{HuJava,Hybrid2016,Python2017,PureScript2019,LinearDecomp}:
these proposals generate APIs in the target language
which the developer
can use to implement their endpoint application
and enjoy the following guarantees:

@@

\subsubsection{API Generation in PureScript \cite{PureScript2019}}
King et al. presented an approach
for integrating \textit{multiparty} session types 
into web development using 
the PureScript language \cite{PureScript}.
PureScript is a strongly typed functional language
that compiles to JavaScript.
This proposal takes advantage of its expressive type system to 
provide \textit{static} guarantees for protocol conformance.
We outline the main components of their EFSM encoding:

\subparagraph{Actions as type classes} 
The semantics of the state transition function 
in the EFSM express that a \texttt{(state, action)} tuple
\textit{uniquely defines} a successor \texttt{state}. 
These semantics can be expressed by 
\textit{multi-parameter type classes (MPTC) 
with functional dependencies}: 
\lstonelineps{class Send r s t a | s -> t r a} 
defines \texttt{Send} as a MPTC 
parameterised by recipient \texttt{r}, 
current state \texttt{s}, successor state \texttt{t} 
and payload type \texttt{a}; 
\texttt{s -> t r a} expresses the functional dependency that, 
for an instance of this type class, 
the current state uniquely determines the
successor state, the recipient and payload type. 
These type classes are independent of the EFSM.

@@

\subsubsection{Session-Typed GUI Programming 
using Links \cite{MVU2020}}
Fowler presented the first formal integration
of session typing and GUI programming.
His work formalised the \textit{Model-View-Update} (MVU)
architectural pattern for GUI development.
An MVU application features a \textit{model} encapsulating application
state, a \textit{view function} rendering the state on the 
user interface, and an
\textit{update function} handling
\textit{messages} produced by the
rendered model to produce a new model.
Fowler introduces the concept of model types
to express type dependencies between these components:
a \emph{model type} uniquely defines a \textit{view} function,
set of \textit{messages} and \textit{update function} --
rather than producing a new model, the update function 
defines valid transitions to other model types.
This approach guarantees channel linearity for GUI programming:
the update function transitions to a successor model, which only
renders an interface with the set of messages unique to that model,
so channel actions from the predecessor model are not rendered,
which makes reuse impossible.
The core MVU calculus is also extended with \textit{commands}
and \textit{linearity} required to express communication
actions in session type theory.

Fowler's proposal \cite{MVU2020} is implemented using
the \textit{Links} web programming language
\cite{LINKS}. Links is a strict, typed, functional language
that can be used to write full-stack web applications;
and similar to PureScript, it is compiled to JavaScript.
As it has a linear type system, session types can be
implemented using native language features to
monitor linear channel usage.
Links also supports error handling for session types,
as introduced in \cite{Exceptional}; 
given side-effects must be made explicit
in functional programming,
Fowler et al. \cite{Exceptional} introduces
a \textit{session cancellation} operator to ensure
that channel resources are closed after handling
side-effects, such as errors.

### 2020-06-18 report/sections/background/sessiontypes.tex
Report: final proofread

The process algebra $\pi$-calculus 
introduced by Milner in \cite{Milner1999} 
formalises the message passing abstraction in terms of 
the basic building blocks of sending and receiving processes.
The composition of these primitives allow us to 
describe more complex communication sessions.
\textit{Session types} define the typing discipline 
for the $\pi$-calculus and provide reliability guarantees 
for communication sessions; 
the latter addresses a key challenge when 
reasoning about the correctness of distributed systems.

Practical application of session types in software
engineering range from
developing languages providing native session type support 
\cite{ATS2016} to implementing session types 
in existing programming languages across different paradigms.
Implementations of the latter approach differ by 
how they leverage the design philosophy and features 
provided by the programming language.
For example, King et al. \cite{PureScript2019} leveraged the 
expressive type system of PureScript to perform 
\textit{static} session type checking,
whilst Neykova and Yoshida \cite{Python2017} 
introduced dynamic approaches to 
check the conformance of Python programs 
with respect to session types.
We discuss these related work, among others, in \cref{section:bgrelated}.

@@

The $\pi$-calculus models concurrent computation, 
where processes can execute in parallel and communicate 
via shared names.
We first consider the \textit{asynchronous} 
$\pi$-calculus introduced by 
Honda and Tokoro in \cite{AsyncHonda}.
Among the many flavours of the calculus 
which vary depending on the application domain,
we outline the variant as presented in \cite{C406Lecture}.

@@

A \textit{binary session} is a parallel composition 
of two processes, each representing a distinct participant.
In the context of web applications, a binary session may
describe the interactions between client and server.
Without loss of generality, a \text{session} represents 
the sequence of send and receive actions of a single participant.

@@

\item \textbf{Labelled messages}: 
A label identifier is attached to all messages; 
the input process in \cref{fig:async}
is generalised as a branching process that offers one branch.
\end{itemize}

@@

\begin{prooftree}
\AxiomC{$\exists	j \in I. l_j = l$}
\AxiomC{$\vec x_j = x_1, \dots, x_n$}
\AxiomC{$e_1 \downarrow v_1 \dots e_n \downarrow v_n$}
\AxiomC{$\mrole{p} \neq \mrole{q}$}
\RightLabel{\rulename{Comm}}
\QuaternaryInfC{$
\mrole{p} :: \sel{q}{l}{e_1 \dots e_n}.~P \mid 
\mrole{q} :: \bra{p}{l_i(\vec x_i): Q_i}{i\in I} 
~\longrightarrow~ 
\ptprocess{p}{P} \mid \ptprocess{q}{Q_j[v_k/x_k]^n_{k=1}}
$}
\end{prooftree}

@@

We derive the type of a process with a {typing judgement} 
of the form $\Gamma \vdash P: S$, which reads, 
\textit{under the typing context $\Gamma$, 
process $P$ has session type $S$}.

The \textit{typing context} records typing assumptions 
used as part of the derivation: in the case of binary session types, 
the context maps expressions to sorts, 
and process variables to session types. 
A typing judgement is constructed in terms of inference rules 
defined inductively on the structure of 
processes and expressions.

@@

\[
\begin{array}{rl}
T_{\text{Client}} &= 
	\tselone{Server}{\tmsg{Succ(nat)}}.~
	\tbraone{Server}{\tmsg{Succ(int)}}.~
	\tend \\
T_{\text{Server}} &= \tbraone{Client}{
\begin{cases}
	\tmsg{Succ(int)}: \tselone{Client}{\tmsg{Succ(int)}}. \tend \\
	\tmsg{Quit()}:  \tend \\
\end{cases}
} \\
\end{array}
\]

Whilst $\dual{T_{\text{Client}}} \neq T_{\text{Server}}$, 
this pair of session types is intuitively compatible 
as the client is selecting a branch offered by the server, 
where the session types for the continuations of 
this branch for both participants are indeed dual.
Regarding payload, the server is expecting \texttt{int},
but the client sends \texttt{nat}, which is a subtype of \texttt{int}.

This motivates the concept of \textit{subtypes} in session types,
which allows a process to be typed by its ``supertype'' when required. 
$\subtype$\footnote{
The $\subtype$ operator is also an 
overloaded relation on 
sorts to express subsorting, 
i.e. $\texttt{nat} \subtype \texttt{int}$.
} defines the subtyping relation: 
$T \subtype T'$ reads \textit{$T$ is a subtype of $T'$},
and is defined coinductively on the structure of $T$.

@@

	& 
	\text{where } T''_k =
	\begin{cases}
	T_k & \text{if } k \in (I \setminus J) \\
	T'_k & \text{if } k \in (J \setminus I) \\
	\tmerge{T_k}{T'_k} & \text{otherwise} \\
	\end{cases}
	& \\

>>>>

The process algebra $\pi$-calculus 
introduced by Milner \cite{Milner1999} 
formalises the message passing abstraction in terms of 
the basic building blocks of sending and receiving processes.
The composition of these primitives allow us to 
describe more complex communication sessions.
\textit{Session types} define the typing discipline 
for the $\pi$-calculus and provide reliability guarantees 
for communication sessions; 
the latter addresses a key challenge when 
reasoning about the correctness of distributed systems.

Practical application of session types in software
engineering range from
developing languages providing native session type support 
\cite{ATS2016} to implementing session types 
in existing programming languages across different paradigms.
Implementations of the latter approach differ by 
how they leverage the design philosophy and features 
provided by the programming language.
For example, King et al. \cite{PureScript2019} leveraged the 
expressive type system of PureScript to perform 
\textit{static} session type checking,
whilst Neykova \cite{Python2017} 
introduced dynamic approaches to 
check the conformance of Python programs 
with respect to session types.
We discuss these related work, among others, 
in \cref{section:bgrelated}.

@@

The $\pi$-calculus models concurrent computation, 
where processes can execute in parallel and communicate 
via shared names.
We first consider the \textit{asynchronous} 
$\pi$-calculus introduced by 
Honda and Tokoro \cite{AsyncHonda}.
Among the many flavours of the calculus 
which vary depending on the application domain,
we outline the variant as presented in \cite{C406Lecture}.

@@

A \textit{session} represents 
the sequence of send and receive actions of 
a single participant. 
A \textit{binary session} describes a session
with two distinct participants.
In the context of web applications, a binary session may
describe the interactions between client and server.

@@

\item \textbf{Labelled messages}: 
A label identifier is attached to all messages; 
the input process in \cref{fig:async}
is generalised as a branching process offering one branch.
\end{itemize}

@@

\begin{prooftree}
\AxiomC{$
\exists j \in I. ~ \left( 
l_j = l
\qquad
\vec x_j = x_1, \dots, x_n 
\qquad
e_1 \downarrow v_1 \dots e_n \downarrow v_n
\right)
$}
\AxiomC{$\mrole{p} \neq \mrole{q}$}
\RightLabel{\rulename{Comm}}
\BinaryInfC{$
\mrole{p} :: \sel{q}{l}{e_1 \dots e_n}.~P \mid 
\mrole{q} :: \bra{p}{l_i(\vec x_i): Q_i}{i\in I} 
~\longrightarrow~ 
\ptprocess{p}{P} \mid \ptprocess{q}{Q_j[v_k/x_k]^n_{k=1}}
$}
\end{prooftree}

@@

We derive the type of a process with a {typing judgement} 
of the form $\Gamma \vdash P: S$, which reads, 
\textit{under the typing context $\Gamma$, 
process $P$ has session type $S$}. 
The \textit{typing context} records typing assumptions 
used as part of the derivation: in the case of binary session types, 
the context maps expressions to sorts, 
and process variables to session types. 
A typing judgement is constructed in terms of inference rules 
defined inductively on the structure of 
processes and expressions.

@@

\[
\begin{array}{rl}
T_{\text{Client}} &= 
	\tselone{Server}{\tmsg{Succ(nat)}}.~
	\tbraone{Server}{\tmsg{Succ(int)}}.~
	\tend \\
T_{\text{Server}} &= \tbraone{Client}{
\begin{cases}
	\tmsg{Succ(int)} &: \tselone{Client}{\tmsg{Succ(int)}}.~\tend \\
	\tmsg{Quit()} &:  \tend \\
\end{cases}
} \\
\end{array}
\]

Whilst $\dual{T_{\text{Client}}} \neq T_{\text{Server}}$, 
this pair of session types is intuitively compatible 
as the client is selecting a branch offered by the server, 
where the session types for the continuations of 
this branch for both participants are indeed dual.
Regarding payload, the server is expecting \texttt{int},
but the client sends \texttt{nat},
which is a \textit{subtype} of \texttt{int}.

This motivates the concept of subtyping in session types,
which allows a process to be typed by its ``supertype'' when required. 
$\subtype$\footnote{
The $\subtype$ operator is also an 
overloaded relation on 
sorts to express subsorting, 
i.e. $\texttt{nat} \subtype \texttt{int}$.
} defines the subtyping relation: 
$T \subtype T'$ reads \textit{$T$ is a subtype of $T'$},
and is defined coinductively on the structure of $T$.

@@

	& 
	\text{where } T''_k =
	\begin{cases}
	T_k & \text{if } k \in I \setminus J \\
	T'_k & \text{if } k \in J \setminus I \\
	\tmerge{T_k}{T'_k} & \text{if } ~ k \in I \cap J \\
	\end{cases}
	& \\
	& \text{otherwise undefined} &\\

### 2020-06-18 report/sections/background/typescript.tex
Report: feedback
Report: final proofread

We introduce the TypeScript language 
\cite{UnderstandingTypeScript}
as our choice of target language
for session type API generation.
Developed by \textit{Microsoft Research},
TypeScript is an extension to
JavaScript to address
the deficiencies of the latter in
\textit{developing} and \textit{maintaining}
large-scale complex applications.
Syntactically,
TypeScript is a \textit{superset} of JavaScript,
so every JavaScript program is a
TypeScript program.
The TypeScript Compiler is used
to compile a TypeScript program into JavaScript
source code; since
JavaScript is a dynamically typed langauge,
the compilation process features full type erasure.

@@

declare function area(sq: ThisSquare): number;

>>>>

We introduce the TypeScript language 
\cite{UnderstandingTypeScript}
as our choice of target language
for session type API generation.
Developed by \textit{Microsoft Research},
TypeScript is an extension to
JavaScript to address
the deficiencies of the latter in
\textit{developing} and \textit{maintaining}
large-scale complex applications.
Syntactically,
TypeScript is a \textit{superset} of JavaScript,
so every JavaScript program is a
TypeScript program.
The TypeScript Compiler is used
to compile a TypeScript program into JavaScript
source code, with full type erasure.

@@

const area = (sq: ThisSquare) => sq.side * sq.side;

### 2020-06-18 report/sections/evalconcl/concl/contributions.tex
Report: feedback on benchmarks
Report: final proofread

We motivate our API generation approach from 
\cite{Hybrid2016,PureScript2019} to generate TypeScript APIs
from a Scribble protocol specification.
Developers write their full-stack applications
by implementing the generated APIs, and enjoy 
communication safety guarantees by construction.
APIs generated for server-side endpoints
are compatible with the Node.js runtime,
as detailed in \cref{chap:node}.
For browser-side endpoints,
we present a novel approach for integrating
session types into web-based GUI programming based on
translating the theory on model types \cite{MVU2020}
to idiomatic practices on the React.js framework;
this is highlighted in \cref{chap:react}.
With respect to session type theory, implementations using
the generated APIs statically enjoy linear channel usage
guarantees and affine channel usage guarantees for back-end
and front-end targets respectively.

Compared to previous work 
\cite{Exceptional,PureScript2019,MVU2020,LINKS} on
session-typed web development, 
we are not only able to
statically provide the same level of communication safety
guarantees from multiparty session type theory, but we do so
using modern web programming practices to increase
the relevance and usability of our work in industry:
our work targets TypeScript, Node.js and React.js,
and we explain in \cref{chap:ext} how we support advanced
web development idioms (such as asynchronous implementations)
and handle premature disconnections gracefully.

We also do not limit support to protocols that describe
a server-centric topology: we formalise 
a theory of routed multiparty session types in 
\cref{chap:theory} to prove that it is possible to relax the
server-centric topology assumption over WebSocket transport in a way 
that preserves communication along with the communication safety 
properties inherited from canonical session type theory.

% PLACES
Our work has received positive feedback from academia.
The intitial part of this project was published as 
\emph{Generating Interactive WebSocket 
Applications in TypeScript} in Volume 314 of the Electronic Proceedings
in Theoretical Computer Science (EPTCS) \cite{PLACES2020},
in which I am the first and lead author.
The materials in \cref{chap:codegen,chap:node,chap:react}
are expanded from \cite{PLACES2020}.
We were also invited to present a part of this work
at the 12th International Workshop on
Programming Language Approaches to Concurrency- \& 
Communication-cEntric Software (PLACES 2020).

>>>>

We motivate our API generation approach from 
\cite{Hybrid2016,PureScript2019} to generate TypeScript APIs
from a Scribble protocol specification.
By writing their full-stack applications using
the generated APIs, developers enjoy 
communication safety guarantees in their endpoint
applications by construction.
APIs generated for server-side endpoints
are compatible with the Node.js runtime,
as detailed in \cref{chap:node}.
For browser-side endpoints,
we present a novel approach for integrating
session types into web-based GUI programming based on
translating the theory on model types \cite{MVU2020}
to idiomatic practices on the React.js framework;
this is highlighted in \cref{chap:react}.
With respect to session type theory, implementations using
the generated APIs statically enjoy linear channel usage
guarantees and affine channel usage guarantees for back-end
and front-end targets respectively.

Compared to previous work 
\cite{Exceptional,PureScript2019,MVU2020,LINKS} on
session-typed web development, 
we are not only able to
statically provide the same level of communication safety
guarantees, but we do so
using modern web programming practices to increase
the relevance and usability of our work in industry:
our work targets TypeScript, Node.js and React.js,
and we explain in \cref{chap:ext} how we support advanced
web development idioms (such as asynchronous implementations)
and handle premature disconnections gracefully.

We also do not limit support to protocols that describe
a server-centric topology: we formalise \newtheory,
a theory of routed multiparty session types, in 
\cref{chap:theory} to prove that it is possible to relax the
server-centric topology assumption over WebSocket transport in a way 
that preserves communication along with the communication safety 
properties inherited from canonical session type theory.

% PLACES
Our work has received positive feedback from academia.
The initial part of this project was published as 
\emph{Generating Interactive WebSocket 
Applications in TypeScript} in Volume 314 of the Electronic Proceedings
in Theoretical Computer Science (EPTCS) \cite{PLACES2020},
in which I am the first and lead author.
The materials in \cref{chap:codegen,chap:node,chap:react}
are expanded from \cite{PLACES2020}.
The materials in 
\cref{chap:ext,chap:p2p,chap:theory,chap:impl,chap:eval}
are developed exclusively in this project by myself.
We were also invited to present a part of this work
at the 12th International Workshop on
Programming Language Approaches to Concurrency- \& 
Communication-cEntric Software (PLACES 2020).

### 2020-06-18 report/sections/evalconcl/concl/future.tex
Report: review chapter 3
Report: final proofread

\cite{ExplicitConnections} uses a variant of the Scribble protocol
language that supports explicit connection actions. Future work
can use that as a basis for code generation, and generalise
our existing mechanisms for session connection and cancellation
to support explicit connection actions.

\paragraph{Transport Abstraction}
The MPST-based web development framework presented by 
King et al. \cite{PureScript2019} parameterised
the transport mechanism in the session runtime definition.
The developer can run the session over different transportation,
provided that the custom transport implements the behaviour
of the \texttt{Transport} type class.

@@

\paragraph{EFSM Encoding as Typestates}
The concept of \textit{typestates} define the interface of
an object to be dependent on its private state, meaning
it can change at runtime.
This is compatible with the EFSM abstraction from the
MPST framework: the permitted methods (i.e. channel actions)
of the EFSM depends on its current state.
In fact, Kouzapas et al. \cite{StMungo} presents a way to
generate typestate specifications for each endpoint of a
Scribble protocol; Gay et al. \cite{ModularST} proposes
an encoding of session types in object-oriented languages
and discuss their approach with respect to typestates.

>>>>

The work of Hu and Yoshida \cite{ExplicitConnections} 
uses a variant of the Scribble protocol
language that supports explicit connection actions. Future work
can use that as a basis for code generation, and generalise
our existing mechanisms for session connection and cancellation
to support explicit connection actions.

\paragraph{Transport Abstraction}
The MPST-based web development framework presented by 
King et al. \cite{PureScript2019} parameterises
the transport mechanism in the session runtime definition.
The developer can run the session over different transportation,
provided that the custom transport implements the behaviour
of the \texttt{Transport} type class.

@@

\paragraph{EFSM Encoding as Typestates}
The concept of \textit{typestates} define the interface of
an object to be dependent on its private state, meaning
it can change at runtime.
This is compatible with the EFSM abstraction from the
MPST framework: the permitted methods (i.e. channel actions)
of the EFSM depend on its current state.
In fact, Kouzapas et al. \cite{StMungo} presents a way to
generate typestate specifications for each endpoint of a
Scribble protocol; Gay et al. \cite{ModularST} proposes
an encoding of session types in object-oriented languages
and discusses their approach with respect to typestates.

### 2020-06-18 report/sections/evalconcl/eval.tex
Report: final proofread

\subsubsection{Expressiveness}
To evaluate the expressiveness
of our work,
we use two case studies
of protocols found in webservices
to demonstrate strengths and weaknesses
of our work:

>>>>

\subsubsection{Expressiveness}
To evaluate the expressiveness
of our work,
we use two case studies
of protocols found in web services
to demonstrate strengths and weaknesses
of our work:

### 2020-06-18 report/sections/evalconcl/eval/benchmarks.tex
Report: final proofread

\begin{figure}[!ht]
\centering
\includegraphics[width=0.5\textwidth]{pingpongclient}
\captionof{figure}{User interface of \tprotocol{Ping Pong} \trole{Client}}
\label{fig:pingpongclient}
\end{figure}

@@

\item We use the production build generated by 
\textit{create-react-app} \cite{cra} for all experiments, which performs the
transpilation into JavaScript. We serve the production build using the
\textit{serve} package \cite{npmserve} available on \textit{npm}.

@@

\item We use the transpiled JavaScript versions of all \trole{Svr}s
for the experiments.

@@

\subsection{Overhead}
The execution time is the total time taken
for the \trole{Client} to connect to the \trole{Server}
and complete the parameterised number of round trips.
We compare the total execution time (\textit{Exec. Time}) 
and execution time per round trip (\textit{Exec. Time / Ping-Pong}) -- 
averaged over 20 repeated experiments -- across the three implementations,
for $n \in \{10^2, 10^3, 10^4\}$.
We summarise the results in \cref{table:overhead}.

@@

render() {
	return this.state.visible 
		? <button
		   ref={this.state.button}
		   onClick={this.click.bind(this)}
		   >Ping</button>
		: <button>Ping</button>;
}
\end{lstlisting}
\captionof{lstlisting}{Preventing user-triggered linearity violation in
\texttt{bare_safe} \tprotocol{Ping Pong}}
\label{lst:workaround}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\begin{subfigure}[b]{\textwidth}
\centering
\begin{lstlisting}[language=javascript,tabsize=2]
socket.onmessage = ({ data }) => {
	const { label, payload } = JSON.parse(data.toString());
	if (label === 'PING') {
		let count: number = payload[0];
		console.timeLog(LABEL, ++count);
		if (count === MSGS) {
			socket.send(JSON.stringify({
				label: 'BYE',
				payload: [count],
			}));
		} else {
			socket.send(JSON.stringify({
				label: 'PONG',
				payload: [count],
			}));
		}
	} else { throw new Error(`Unrecognised label: ${label}`); }
}
\end{lstlisting}
\caption{\texttt{bare} implementation of \tprotocol{Ping Pong} \trole{Svr}}
\label{lst:nodepingpongbare}
\end{subfigure}
\hfill
\begin{subfigure}[b]{\textwidth}
\centering
\begin{lstlisting}[language=javascript,tabsize=2]
const logic = new Implementat.S14({
	PING: (count) => {
		console.timeLog(LABEL, ++count);
		if (count === MSGS) {
			return new Implementation.S16([
				Labels.S16.BYE, [count], new Implementation.S15()
			]);
		} else {
			return new Implementation.S16([
				Labels.S16.PONG, [count], logic
			]);
		}
 	}
});
\end{lstlisting}
\caption{\texttt{mpst} implementation of \tprotocol{Ping Pong} \trole{Svr}}
\label{lst:nodepingpongmpst}
\end{subfigure}
\caption{Comparison of \texttt{bare} and \texttt{mpst}
implementations for \tprotocol{Ping Pong} \trole{Svr}}
\label{fig:nodepingpong}
\end{figure}

>>>>

\begin{figure}[!ht]
\centering
\includegraphics[width=0.5\textwidth]{pingpongclient}
\captionof{figure}{User Interface of \trole{Client} Endpoint in
\tprotocol{Ping Pong} Protocol}
\label{fig:pingpongclient}
\end{figure}

@@

\item We use the production build generated by 
\textit{create-react-app} \cite{cra} for all experiments, which performs the
compilation into JavaScript. We serve the production build using the
\textit{serve} package \cite{npmserve} available on \textit{npm}.

@@

\item We use the compiled JavaScript versions of all \trole{Svr}s
for the experiments.

@@

\subsection{Overhead}
The execution time is the total time taken
for the \trole{Client} to connect to the \trole{Server}
\textit{and} complete the parameterised number of round trips.
We compare the total execution time (\textit{Exec. Time}) 
and execution time per round trip (\textit{Exec. Time / Ping-Pong}) -- 
averaged over 20 repeated experiments -- across the three implementations,
for $n \in \{10^2, 10^3, 10^4\}$.
We summarise the results in \cref{table:overhead}.

@@

render() {
	return this.state.visible 
		? <button
		   ref={this.state.button}
		   onClick={this.click.bind(this)}
		   >Ping</button>
		: <button>Ping</button>;
}
\end{lstlisting}
\captionof{lstlisting}{Preventing Channel Linearity Violation in
\texttt{bare_safe} \tprotocol{Ping Pong}}
\label{lst:workaround}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\begin{subfigure}[b]{\textwidth}
\centering
\begin{lstlisting}[language=javascript,tabsize=2]
socket.onmessage = ({ data }) => {
	const { label, payload } = JSON.parse(data.toString());
	if (label === 'PING') {
		let count: number = payload[0];
		console.timeLog(LABEL, ++count);
		if (count === MSGS) {
			socket.send(JSON.stringify({
				label: 'BYE',
				payload: [count],
			}));
		} else {
			socket.send(JSON.stringify({
				label: 'PONG',
				payload: [count],
			}));
		}
	} else { throw new Error(`Unrecognised label: ${label}`); }
}
\end{lstlisting}
\caption{\texttt{bare} Implementation of \tprotocol{Ping Pong} \trole{Svr}}
\label{lst:nodepingpongbare}
\end{subfigure}
\hfill
\begin{subfigure}[b]{\textwidth}
\centering
\begin{lstlisting}[language=javascript,tabsize=2]
const logic = new Implementation.S14({
	PING: (count) => {
		console.timeLog(LABEL, ++count);
		if (count === MSGS) {
			return new Implementation.S16([
				Labels.S16.BYE, [count], new Implementation.S15()
			]);
		} else {
			return new Implementation.S16([
				Labels.S16.PONG, [count], logic
			]);
		}
 	}
});
\end{lstlisting}
\caption{\texttt{mpst} Implementation of \tprotocol{Ping Pong} \trole{Svr}}
\label{lst:nodepingpongmpst}
\end{subfigure}
\caption{Comparison of \texttt{bare} and \texttt{mpst}
Implementations for \tprotocol{Ping Pong} \trole{Svr}}
\label{fig:nodepingpong}
\end{figure}
